\documentclass[mathserif,notheorems]{beamer} % option notheorems
\usepackage{amssymb, amsmath, amsthm, amsfonts}
\usepackage{cleveref}
\usepackage{thmtools}
\usetheme{Berlin}
\title{A Review of Lossless Data Compression Algorithms}
\author{Jack Stanek}

\declaretheorem[numberwithin=section]{Theorem}

\begin{document}
\begin{frame}
  \titlepage
\end{frame}

\begin{frame}
  \tableofcontents
\end{frame}

\begin{frame}
  \section{A Brief Introduction to Data Compression}
  \subsection{Informal Overview}
  \frametitle{What is data compression?}

  Informally, a data compression scheme is some method for reducing
  the amount of space required to store a given amount of data.

  Some commonplace examples:

  \begin{itemize}
  \item \texttt{*.mp3} files
  \item \texttt{*.zip} files
  \item Abbreviations in writing
  \end{itemize}

  This brief list includes some lossy methods, though this
  presentation will cover only lossless methods.
\end{frame}

\begin{frame}
  \subsection{Theoretical Background}
  Information can be represented as a series of symbols
  \begin{itemize}
  \item Text is a string of letters in langauge's alphabet
  \item Numbers are strings of Arabic numerals
  \end{itemize}
  For some alphabet $\Sigma$, the set of all compression schemes
  $C_\Sigma$ is the set of bijections
  $c : \Sigma^n \to \bigcup_{k = 1}^{n} \Sigma^k$ for all
  $n \in \mathbb{N}$ such that $n \geq 2$.
\end{frame}

\begin{frame}
  \subsection{Impossibility of Perfect Compression}
  A compression scheme is considered ``perfect'' if and only if its
  image is the set of all strings of length strictly less than that of
  the domain.
  \begin{Theorem}
    No perfect compression scheme exists.
  \end{Theorem}
\end{frame}

\begin{frame}
  \subsection{Maximum Compressibility of Data}
  \begin{itemize}
  \item This result suggests there is a limit to how small compressed
    data can be.
  \item Shannon's Noiseless Coding gives a lower bound.
    \begin{itemize}
    \item The average word size of a prefix code on a channel is
      greater than or equal to its entropy.
    \item Since human data sources are typically very regular
      (i.e. have low information entropy), data compression methods
      are able to greatly reduce the sizes of these sorts of data.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \section{Lempel-Ziv Based Algorithms}
  \subsection{LZ77 and LZ78}
  \frametitle{Lempel-Ziv compression algorithms}
  LZ77 and LZ78 (a.k.a. LZ1 and LZ2, a.k.a. LZ)
  \begin{itemize}
  \item Introduced in 1977 and 1978 by Abraham Lempel and Jakob Ziv
  \item Dictionary block compressor family
  \item Uses a sliding-window approach
  \item Produces 3-tuples representing prefixes
  \end{itemize}
\end{frame}

\begin{frame}
  \subsection{LZ77 Example}
  \frametitle{A Brief Example}

  Suppose we are to encode the word ``hahahahaha'' with a window size
  of 16.
\end{frame}

\begin{frame}
  \frametitle{LZ78}
  LZ78 is a slightly different modification on the LZ77 algorithm.
  \begin{itemize}
  \item Instead of a sliding window, it uses a ``chained dictionary''
  \item Each entry in the dictionary is a two-tuple: \texttt{(
      previous prefix index, following character)}
  \item The dictionary can be represented as a tree data structure
  \item Equivalent to LZ77 even though it uses a different method
  \end{itemize}
\end{frame}

\begin{frame}
  \subsection{LZW and DEFLATE}
  \frametitle{Derivative Methods: LZW and DEFLATE}

  Since they are very effective and efficient algorithms, LZ77 and
  LZ78 have inspired a wide variety of derivative methods.

  \begin{itemize}
  \item LZW (Lempel-Ziv-Welch): A method based on LZ78 which begins
    with a pre-populated prefix dictionary, with each individual
    character in the alphabet.
  \item DEFLATE: Splits the input into blocks, then encodes the
    alphabet using a Huffman code, then compresses the resultant input
    channel using LZ78.
  \end{itemize}
\end{frame}

\begin{frame}
  \section{FLAC}
  \frametitle{Free Lossless Audio Codec}
  The Free Lossless Audio Codec is a domain-specific compression
  scheme for hi-fi audio.
  \begin{enumerate}
  \item Split input into blocks (can optimize for compression or
    speed)
  \item Encode stereo channels (left-right or mid-side)
  \item Fit the sound signal to an n-degree polynomial (configurable)
  \item Calculate the residuals (difference between the raw signal and
    the fit polynomial)
  \item Encode the residual values using Rice codes (essentially a
    Huffman coding optimized for alphabets with geometric
  \end{enumerate}
\end{frame}

\begin{frame}
  \section{Question and Answer}
  \frametitle{Any questions?}
\end{frame}

\begin{frame}
  \frametitle{Thank you!}
\end{frame}
\end{document}
